Verification of Stencil Computations through Spatial Specifications

### 2016-07-05 paper.tex
Consistency and checking nearly explained

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{amsthm}
\usepackage{tikz}
\usepackage{multirow}
\usepackage{minted}

@@

@@

%% VECTOR NOTATIONS
\newcommand{\vect}[1]{\textbf{#1}}
\newcommand{\vtwoh}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{cc}#1 \; & \; #2\end{array}\right]}
\newcommand{\vthreeh}[3]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{ccc}#1 \; & \; #2 \; & \; #3\end{array}\right]}
\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{l}$#1$\\$#2$\end{array}\right]}
\newcommand{\vthree}[3]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{l}$#1$\\$#2$\\$#3$\end{array}\right]}
\newcommand{\stwo}[4]
%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
{\setlength{\arraycolsep}{0.1em}
\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{specTop} \;
                        \texttt{::} \; v
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[1em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
\nonterm{spec} ::= & [\nonterm{multMod},] \; \nonterm{region} \\
\nonterm{multMod} ::= & \term{readOnce} \\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \\
\multicolumn{2}{l}{\qquad\qquad\;\;\; \stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}}  \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+} \; \nonterm{region}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

Figure~\ref{fig:equations} lists the equational theory for
specifications which is defined by the binary relation $\equiv$
between region terms. Stencil declarations are considered
equal when they have the same modifiers and $\equiv$ equates
their regions, \ie{}:
\begin{align*}
& \textit{region} \equiv \textit{region'} \\
& \quad \Rightarrow \;\; \texttt{stencil} \; [\textit{approxMod},] \; [\textit{multMod},] \;
\textit{region} \\[-0.4em]
& \quad\quad \;\; \equiv \;\texttt{stencil} \; [\textit{approxMod},] \;
            [\textit{multMod},] \; \textit{region}'
\end{align*}
%
To save space in the equations of Figure~\ref{fig:equations}, we use abbreviations
\term{refl}, \term{fwd}, \term{bwd}, and \term{cen} for \term{reflexive},
\term{forward}, \term{backward}, and \term{centered} regions
respectively. Futhermore, the abbreviations include syntactic sugar
for quantifying over the abseence or presence of the \irrefl{}
attribute on region constants as a boolean. For example, 
the syntactic sugar in our equations for \term{fwd} is defined:
%%
\begin{align*}
& \stenFwdSR{n}{d}{r} 
\; := \; \\[-0.4em]
& \qquad \begin{cases}
\stenFwd{n}{d}{, \irrefl} & \textit{if} \;\, \neg \, r \\
\stenFwd{n}{d}{}  & \textit{if} \;\,  r
\end{cases} \\
\textit{or} \;\;
& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
\end{align*}
%
and similarly for \term{bwd} and \term{cen}.
We also frequently abbreviate \irrefl{} to \irreflS{}.

@@

We define a semantic model of our stencil specification language as
the basis for building correct implementations, showing the soundness
of the equational theory and approximations, and as a basis for the
inference, checking, and program synthesis algorithms.
The semantics of stencil specifications can be given a denotational
model as sets of vectors, represent array indexing.

\begin{definition}
  A specification of maximum dimensionality $n$, is modelled by a
  set of vectors of size $n$ ($n$-dimensional vectors or 
  $n$-vectors for short)
  with values drawn from $\mathbb{Z} \cup \{\infty\}$, representing
  relative indices. The element $\infty$ represents non-relative
  indices, such as constants, which are seen as infinite with respect
  to the frame of reference of a relative offset from an induction
  variable.

  For example, the array index term $(i, j+1)$ corresponds to the vector
  $\vtwoh{0}{1}$, whereas, the index term $(3, j)$ corresponds to the
  vector $\vtwoh{\infty}{0}$ since $3$ is a non-relative index.
\end{definition}

We define the interpretation function $\interp{-}_n$ where $n$ is the
maximum dimensionality of the specification being modelled, overloaded
on regions in Figure~\ref{fig:region-model} and on the top-level of a
specification \textit{specTop} in Figure~\ref{fig:spatial-model}. Some
intermediate notations and definitions are used.

\begin{definition}A \emph{single-entry vector} of size $n$, denoted
$\singleEntry{r}{n}$, is a vector
where the $r^{th}$ entry is $1$ and all others are $\infty$. For
example, 
$\singleEntry{1}{2} = \vtwoh{\infty}{1}$ and $\singleEntry{2}{3} = 
\vthreeh{1}{\infty}{1}$. 
\end{definition}

\begin{definition}A \emph{zero-entry vector} of size $n$, denoted
$\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{\infty}{0}$.
\end{definition}

The first four equations of Figure~\ref{fig:region-model}
give the meanings of the region constants as sets of indices.

\begin{definition}The product of two models of dimensionality
$n$ is given by the tensor $\otimes_n$ defined:
%%
\begin{alignat*}{2}
& \effdims{M}{n} =
\bigcup_{u \in M} \{x \mid x \in \{1,\ldots,n\}, u_x \neq \infty\} \\
& M \otimes_n N =
  \Bigg\{x \; \Bigg| \;
    \parbox{5.6cm}{$i \in \{1, \ldots, 2^n\},$
                   $(u, v) \in M \times N,$ \\
                   $x = (u \bowtie v)_i,$ \\
                   $j \in (\effdims{M}{n} \cup \effdims{N}{n}),$ \\
                   $x_j \neq \infty$
                  } \Bigg\}
\end{alignat*}
The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
product of $M \times N$.

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-deterministic picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise permutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$. Recall that the vectors are defined over the domain $\mathbb{Z} \cup
\{\infty\}$. In pairwise permutation $\infty$

%%
\end{definition}

@@

\begin{figure}
\input{spatial-model.tex}
\caption{Model of specifications, $\interp{\textit{specTop}}$}
\label{fig:spatial-model}
\end{figure}

\begin{figure}
\input{consistency.tex}
\caption{Consistency function for a model and a set of indices}
\label{fig:consistency}
\end{figure}

@@

\section{Inference, checking, and synthesis}
\label{sec:analysis}

\mnote{Few cases that we might like to consider (there are examples
    of each in \#camfort-main Slack channel):
  \begin{itemize}
    \item Matrix operator overloading $a + b$, where both are arrays
    \item Scalar operator overloading $a + b$, where a is an array b is a
      scalar. $b$ is added to all.
    \item Sliding. $a = b$ where the dimensions are $a(1:10)$, $b(2:11)$,
      so the assignment causes shifting.
  \end{itemize}
}

@@

\subsection{Inference\label{sec:inference}}

@@

@@

We first examined how frequently stencil computations occur in our
corpus. We define a \emph{potential} statement as one whose left-hand
side is an array subscript on neighbourhood indices. These statements
are the inputs to Step 2 of the Inference process described
Section~\ref{sec:inference}. Each potential statement for which we
eventually generate some stencil specifications for is also counted as
an \emph{actual} statement.

>>>>

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{amsthm}
\usepackage{tikz}
\usepackage{multirow}
\usepackage{minted}
\usepackage{breqn}

@@

\newcommand\mdef{\stackrel{\mathclap{\normalfont\mbox{def}}}{=}}

@@

%% VECTOR NOTATIONS
\newcommand{\vect}[1]{\textbf{#1}}
\newcommand{\vtwoh}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{cc}#1 \; & \; #2\end{array}\right]}
\newcommand{\vthreeh}[3]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{ccc}#1 \; & \; #2 \; & \; #3\end{array}\right]}
\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{l}#1\\#2\end{array}\right]}
\newcommand{\vthree}[3]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{l}$#1$\\$#2$\\$#3$\end{array}\right]}
\newcommand{\stwo}[4]
%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
{\setlength{\arraycolsep}{0.1em}
\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[1em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{spec} ::= & [\nonterm{approx},] \; [\nonterm{mult},] \; \nonterm{region} \\
\nonterm{mult} ::= & \term{readOnce} \\
\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \\
\multicolumn{2}{l}{\qquad\qquad\;\;\; \stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}}  \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+} \; \nonterm{region}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

Figure~\ref{fig:equations} lists the equational theory for
specifications which is defined by the binary relation $\equiv$
between region terms. Stencil declarations are considered
equal when they have the same modifiers and $\equiv$ equates
their regions, \ie{}:
\begin{align*}
& \textit{region} \equiv \textit{region'} \\
& \quad \Rightarrow \;\; \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
\textit{region} \\[-0.4em]
& \quad\quad \;\; \equiv \;\texttt{stencil} \; [\textit{approx},] \;
            [\textit{mult},] \; \textit{region}'
\end{align*}
%
To save space in the equations of Figure~\ref{fig:equations}, we use abbreviations
\term{refl}, \term{fwd}, \term{bwd}, and \term{cen} for \term{reflexive},
\term{forward}, \term{backward}, and \term{centered} regions
respectively. Futhermore, the abbreviations include syntactic sugar
for quantifying over the abseence or presence of the \irrefl{}
attribute on region constants as a boolean. For example,
the syntactic sugar in our equations for \term{fwd} is defined:
%%
\begin{align*}
& \stenFwdSR{n}{d}{r}
\; := \; \\[-0.4em]
& \qquad \begin{cases}
\stenFwd{n}{d}{, \irrefl} & \textit{if} \;\, \neg \, r \\
\stenFwd{n}{d}{}  & \textit{if} \;\,  r
\end{cases} \\
\textit{or} \;\;
& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
\end{align*}
%
and similarly for \term{bwd} and \term{cen}.
We also frequently abbreviate \irrefl{} to \irreflS{}.

@@

We define a semantic model of our stencil specification language as
the basis for building correct implementations, showing the soundness
of the equational theory and approximations, and as a basis for the
inference, checking, and program synthesis algorithms.
The semantics of stencil specifications can be given a denotational
model as sets of unifiers for array indexing.

\begin{definition}
  A specification of maximum dimensionality $n$, is modelled by a
  set of unifiers. Here, a unifier is represented as a vector of size $n$
  whose values are drawn from $\mathbb{Z} \cup \{\infty\}$. The purpose
  of the unifier is to accept or reject relativised indices of an array
  subscript. Formal definition of unification in this context is given
  later.

  The element $\infty$ represents the general acceptor meaning it would unify
  with any index in the same dimension including constants or any offset from
  an induction variable.
\end{definition}

\begin{definition} Unification relation, $\sim$, establishes whether
  a particular relativised subscript, $u$, of a group of subscripts involved
  in a stencil computation conforms with a unifier, $v$, of the model for
  the stencil computation.

  Formally, the unification relation can be defined as follows:
  $$u \sim_n v \triangleq
  \{ (u,v) \; \mid \; \forall i \in \{ 1, \ldots, n \}, v_i = \infty \vee u_i = v_i \}$$
\end{definition}

For example, consider a unifier of the form $\vtwoh{\infty}{1}$.

An array index term $(i, j+1)$ would corresponds to the vector
$\vtwoh{0}{1}$ after relativisation. This would unify with the
$\vtwoh{\infty}{1}$ as the first dimension of the unifier have a
general acceptor and the second dimension matches the relative index.

Another index term $(3, j)$ corresponds to the vector $\vtwoh{\infty}{0}$
after relativisation since $3$ is a non-relative index. This would fail
to unify. In the first dimension non-relative index would unify with
the general acceptor but in the second dimension 1 wouldn't unify with 3.

We define the interpretation function $\interp{-}_n$ where $n$ is the
maximum dimensionality of the specification being modelled, overloaded
on regions in Figure~\ref{fig:region-model} and on the top-level of a
specification \textit{spec} in Figure~\ref{fig:spatial-model}. Some
intermediate notations and definitions are used.

\begin{definition}A \emph{single-entry vector} of size $n$, denoted
$\singleEntry{r}{n}$, is a vector where the $r^{th}$ entry is $1$
and all others are $\infty$, \eg{}, $\singleEntry{1}{2} =
\vtwoh{1}{\infty}$ and $\singleEntry{2}{3} = \vthreeh{\infty}{1}{\infty}$.

Here the meaning of a scaled single-entry vector is that at the
dimension it is not $\infty$, it provides a unifier for consistent
relativised array subscript indices. In particular the constrained
requires a non-zero offset from the induction variable.
\end{definition}

\begin{definition}A \emph{zero-entry vector} of size $n$, denoted
$\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.

Similar to the single-entry vector, zero-entry vector also imposes
a constaint on relativised indices. The only difference is the indexing
expression crosses the origin with respect to the induction variable.
\end{definition}

The first four equations of Figure~\ref{fig:region-model}
give the meanings of the region constants as sets of unifiers.
The next two shows how regions are composed in the model.

\begin{definition}The product of two models of dimensionality
$n$ is given by the tensor $\otimes_n$ defined:
%%
\begin{alignat*}{2}
& \effdims{M}{n} =
\bigcup_{u \in M} \{x \mid x \in \{1,\ldots,n\}, u_x \neq \infty\} \\
& M \otimes_n N =
  \Bigg\{x \; \Bigg| \;
    \parbox{5.6cm}{$i \in \{1, \ldots, 2^n\},$
                   $(u, v) \in M \times N,$ \\
                   $x = (u \bowtie v)_i,$ \\
                   $j \in (\effdims{M}{n} \cup \effdims{N}{n}),$ \\
                   $x_j \neq \infty$
                  } \Bigg\}
\end{alignat*}
In plain language, tensor product combines non-$\infty$ constraints
of the unifiers in different dimensions into less accepting unifiers.
This means an access pattern unifying with this unifier would require
both models to be satisfied. Use of $dims$ gives the dimensions that
constitutent models talk about and is used in the product to eliminate
general unifier appearing in these dimensions. Allowing general unifier
in these dimensions would prevent both models to be satisfied
simultaneously.

The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
product of $M \times N$.

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-deterministic picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise permutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$.
%%
\end{definition}

@@

\begin{figure}
\input{spatial-model.tex}
\caption{Model of specifications, $\interp{\textit{spec}}$}
\label{fig:spatial-model}
\end{figure}

@@

\section{Inference and checking}
\label{sec:analysis}

@@

\subsection{Inference\label{subsec:inference}}

@@

\subsection{Checking\label{subsec:checking}}

Verification occurs when a user supplied specification is compared to the
actual access pattern. The mathematical model developed in
Section~\ref{sec:semantics} is used this comparison.

A specification is first synthesised into its model. Then access patterns
are checked against the model for consistency as defined in
Figure~\ref{fig:consistency}. The consistency predicate is defined recursively.
It first checks multiplicity if $\textsf{once}$ is at the top level, \ie{}
specification in question involves \texttt{readOnce} modifier, then array
accesses are checked for uniqueness using $unique$ predicate and the problem
is reduced to $\textsf{mult}$, which does not impose any extra constraints.
Next, consistency predicate considers whether the stencil is an approximation.
Semantically, a lower bound is achieved if every unifier in the model
correspond to some relativised vector of indices as that means all cells
described in the specification are part of the access pattern. Conversely,
an upper bound means all observed access patterns have a unifier. Hence,
the actual cells are a subset of the specified ones. In combination an
exact specification, \ie{} one without any approximation modifiers, is
one that is both a lower and an upper bound.

\begin{figure}
\input{consistency.tex}
\caption{Consistency function of a model against stencil computation}
\label{fig:consistency}
\end{figure}

We will consider both a consistent and an inconsistent example to show how
checking works. The following is an example with a looser but still consistent
version of the \textit{fivepoint} stencil specification inferred previously in
the Subsection~\ref{subsec:inference}:
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j)
\end{minted}
%
\begin{enumerate}
\item Generate model from the specification as described in
  Section~\ref{sec:semantics}.
%
\begin{dmath*}
\interp{\texttt{stencil readOnce, fivepoint}}_2 =
  \textsf{once}(
    \textsf{exact}(
      \{ \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
          \vtwo{\infty}{1}\}))
\end{dmath*}
%
\item Now $cons$ function is applied with this model along side with the
  assignment on line three and array $b$. For brevity the assignment is
  referred as $L3$ and the set of unifiers defined in the previous step as
  $m$.

  The predicate is reduced to
  $cons(\textsf{once}(\textsf{exact}(m)),L3,b)$.
%
\item $cons$ predicate from the previous step evaluates to
  $$unique(L3,b) \; \wedge \; cons(\textsf{mult}(\textsf{exact}(m)),L3,b).$$
%
  The former condition holds as no access pattern is repeated in $L3$ by simple
  inspection.
%
\item We further expand the function to 
  \begin{equation*}
    cons(\textsf{mult}(\textsf{up}(m)),L3,b) \wedge 
    cons(\textsf{mult}(\textsf{low}(m),L3,b).
  \end{equation*}
%
\item To evaluate the two predicates above, we need relativised indices for the
  array in question. This is obtained by $analyse$ function from
  Subsection~\ref{subsec:inference}.

  For the example above the relativised indices are as follows:
%
\begin{dmath*}
analyse(L3,b) =
  \{ \vtwo{0}{-1},
     \vtwo{0}{0},
     \vtwo{0}{1},
     \vtwo{-1}{0},
     \vtwo{1}{0}
  \}
\end{dmath*}
%
\item Below we present two match ups of relativised access patterns and unifiers
  to show both the lower and the upper bound holds.

  The unifications for the lower bound:
  \begin{align*}
     \vtwoh{0}{-1} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{0}{0} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{0}{1} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{-1}{0} & \sim_2 \vtwoh{\infty}{0} \\
     \vtwoh{1}{0} & \sim_2 \vtwoh{\infty}{0}
  \end{align*}

  The unifications for the upper bound:
  \begin{align*}
     \vtwoh{-1}{0} & \sim_2 \vtwoh{-1}{\infty} \\
     \vtwoh{0}{0} & \sim_2 \vtwoh{0}{\infty} \\
     \vtwoh{1}{0} & \sim_2 \vtwoh{1}{\infty}  \\
     \vtwoh{0}{-1} & \sim_2 \vtwoh{\infty}{-1} \\
     \vtwoh{0}{0} & \sim_2 \vtwoh{\infty}{0} \\
     \vtwoh{0}{1} & \sim_2 \vtwoh{\infty}{1}
  \end{align*}

  Since both predicates are satisfied the array accesses specified on line three
  are consistent with the \texttt{fivepoint} specification.
\end{enumerate}

@@

We first examined how frequently stencil computations occur in our
corpus. We define a \emph{potential} statement as one whose left-hand
side is an array subscript on neighbourhood indices. These statements
are the inputs to Step 2 of the Inference process described
Section~\ref{subsec:inference}. Each potential statement for which we
eventually generate some stencil specifications for is also counted as
an \emph{actual} statement.
